---
title: "Class 12"
author: "Karleen Guerrero"
date: "2024-11-11"
output: pdf_document
---

>Q13. Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding.

```{r}
data <- read.table("rs8067378_ENSG00000172057.6.txt", header = TRUE)

summary(data)

library(dplyr)

summary_data <- data %>%
  group_by(geno) %>%
  summarise(
    Sample_Size = n(),
    Median_Expression = median(exp, na.rm = TRUE)
  )

print(summary_data)

```


>Q14 Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.

```{r}
data <- read.table("rs8067378_ENSG00000172057.6.txt", header = TRUE, sep = " ", na.strings = c("", "NA"))
str(data)
head(data)
summary(data)
colnames(data) <- c("sample", "geno", "exp")

# Filter out rows with NA 
data_clean <- data[!is.na(data$exp) & is.finite(data$exp), ]

# boxplot 
boxplot(exp ~ geno, data = data_clean,
        main = "Expression Levels by Genotype",
        xlab = "Genotype", ylab = "Expression Levels",
        col = c("blue", "orange", "purple"))
```



