---
title: "Class 13"
author: "Karleen Guerrero (16791042)"
format: html
---

##Import/Read the data from Himes et. al

```{r}
counts <- read.csv("class 13.csv", row.names=1)
meta <- read.csv("class 13..csv", row.names=1)
```


Lets have a wee peak at this data 
```{r}
head(metadata)
```






































I need to translate our gene identifiers "ENSG0000..." into genes names that the rest of the world can understand.
To do this annotation I will use the "AnnotationDBi" package. I can install this with BiocManager::install()'

```{r}
library(AnnotationDBi)
library(org.Hs.eg.db)

columns(org.Hs.eg.db)
```
I will use the 'mapIds()' function to "Map" my identifiers to those from different databases. will go between "ENSEMBL" and SYMBOL" (and then after "GENENAME")

```{r}
res$symbol <- mapIds(org.Hs.eg.db, 
                     keys = rownames(res), 
                     keytype = "ENSEMBL", 
                     column = "SYMBOL")

#head(res)
```

Add "GENENAME"
```{r}
res$genename <- mapIds(org.Hs.eg.db, 
                     keys = rownames(res), 
                     keytype = "ENSEMBL", 
                     column = "SYMBOL")
```
Add "ENTREZID"
```{r}
Res$entrez <- mapIds(org.Hs.eg.db, 
                     keys = rownames(res), 
                     keytype = "ENSEMBL", 
                     column = "ENTREZID")
```

Save out annotated results object
```{r}
write.csv(res, file="results_annotated.csv")
```
# Pathway Analysis 

Now that we have out results with added annotation we can do some pathway mapping. 
Lets use the **gage** package to loo for KEGG pathways in our results (genes of interest)

```{r}
#| message: false
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 1)
```


What **gage** wants as inpuut is not my big table/data.frame of results. it just want a "vector of importance"/ For RNAseq data like we have this is out log2FC values...

```{r}
x <- c(10, 20, 100)
names(x) <- c(barry)
```


```{r}
foldchanges = res$log2FoldChange
names(foldchanges) = res$entrez
head(foldchanges)
```

Now, lets run the gage pathway analysis. 
```{r}
# Get the results
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

What is in this kgreggs? 


```{r}

```

```{r}
head(keggres$less, 3)
```

Let's use the pathview package to look at one of these highlighted KEGG pathways with out genes highlighted. "hsa05310 Asthma"

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```



```{r}
! [Asthma Pathway with my DEGs](hsa05310.pathway.png)
```

